---
title: "Bayesian Model Analysis"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

# 1. Overview

This section implements the Bayesian regression model for used-car prices. 
Exploratory analysis and initial data cleaning were performed separately.

Goal: 

-   estimate the effect of vehicle characteristics on used-car price

-   quantify uncertainty in these effects using posterior distributions

-   generate predictive distributions for car prices

We model **log(price)** because raw prices are right-skewed.


# 2. Load packages 

```{r }
library(tidyverse)
library(janitor)
library(forcats)
library(MASS)
library(coda)
```


# 3. Load and prepare data

```{r }
cars <- read_csv("../data/train.csv") %>%
  clean_names()
```

We start from the cleaned dataset used in the exploratory analysis.

## 3.1 Restrict to Major Brands
```{r }
top_brands <- c(
  "BMW","Ford","Mercedes-Benz","Chevrolet","Audi","Porsche",
  "Toyota","Lexus","Jeep","Land","Cadillac","Nissan"
)

cars_top <- cars %>%
  filter(brand %in% top_brands)
```

## 3.2 Create Modeling Variables

We construct simplified variables for regression.
```{r }
cars_model <- cars_top %>%
  mutate(
    log_price = log(price),

    accident_bin = if_else(grepl("None", accident),0,1),

    trans_auto = if_else(
      grepl("Automatic|A/T|CVT|Dual", transmission, ignore.case = TRUE),
      1,0
    ),

    milage_std = as.numeric(scale(milage)),
    model_year_std = as.numeric(scale(model_year))
  )
```

Continuous predictors are standardized to improve numerical stability.

## 3.3 Remove Missing Values

For simplicity, we use complete-case analysis. 
```{r }
cars_model <- cars_model %>%
  drop_na(
    log_price,
    milage_std,
    model_year_std,
    accident_bin,
    trans_auto,
    fuel_type,
    brand
  )
```


# 4. Construct the Design Matrix

We build the regression design matrix using selected predictors. 
```{r }
X <- model.matrix(
  ~ milage_std + model_year_std + accident_bin + trans_auto + fuel_type + brand,
  data = cars_model
)

y <- cars_model$log_price

n <- nrow(X)
p <- ncol(X)
```


# 5. Bayesian Model Setup 

Model: 
$$
y \sim N(X\beta,\sigma^2)
$$
Priors:

-   $\beta \sim N(0,\tau^2 I)$

-   $\sigma^2 \sim IG(a,b)$ 

```{r }
tau2 <- 100^2
a0 <- 0.01
b0 <- 0.01
```

These are weakly informative priors.


# 6. Gibbs Sampling

We implement Gibbs sampling using the conjugate full conditional distributions.
```{r }
set.seed(724)

n_iter <- 12000
burn <- 2000

beta_draws <- matrix(NA,n_iter,p)
sigma2_draws <- numeric(n_iter)

beta <- rep(0,p)
sigma2 <- var(y)

XtX <- t(X)%*%X
Xty <- t(X)%*%y
Ip <- diag(p)

for(i in 1:n_iter){

  Vb <- solve(XtX/sigma2 + Ip/tau2)
  mb <- Vb %*% (Xty/sigma2)

  beta <- as.numeric(MASS::mvrnorm(1,mb,Vb))

  resid <- y - X%*%beta

  a_n <- a0 + n/2
  b_n <- b0 + sum(resid^2)/2

  sigma2 <- 1/rgamma(1,a_n,b_n)

  beta_draws[i,] <- beta
  sigma2_draws[i] <- sigma2
}

beta_post <- beta_draws[(burn+1):n_iter,]
sigma2_post <- sigma2_draws[(burn+1):n_iter]

colnames(beta_post) <- colnames(X)
```


# 7. MCMC Diagnostics

We use trace plots and posterior density plots to examine convergence and mixing.

## 7.1 Prepare posterior samples for plotting
```{r }
beta_df <- as.data.frame(beta_post)
beta_df$iteration <- 1:nrow(beta_df)

beta_long <- beta_df %>%
  pivot_longer(
    cols = -iteration,
    names_to = "parameter",
    values_to = "value"
  ) %>%
  mutate(
    parameter = recode(
      parameter,
      "(Intercept)" = "Intercept",
      "milage_std" = "Mileage (standardized)",
      "model_year_std" = "Model Year (standardized)",
      "accident_bin" = "Accident History",
      "trans_auto" = "Automatic Transmission",
      "fuel_typeDiesel" = "Fuel: Diesel",
      "fuel_typeElectric" = "Fuel: Electric",
      "fuel_typeGasoline" = "Fuel: Gasoline",
      "fuel_typeHybrid" = "Fuel: Hybrid",
      "fuel_typePlug-In Hybrid" = "Fuel: Plug-In Hybrid"
    )
  )
```

## 7.2 Trace plots for selected coefficients

We inspect trace plots for a subset of key regression coefficients.
```{r }
selected_params <- c(
  "Intercept",
  "Mileage (standardized)",
  "Model Year (standardized)",
  "Accident History",
  "Automatic Transmission",
  "Fuel: Diesel"
)

beta_long %>%
  filter(parameter %in% selected_params) %>%
  filter(iteration %% 5 == 0) %>%
  ggplot(aes(x = iteration, y = value)) +
  geom_line(linewidth = 0.3, alpha = 0.7) +
  facet_wrap(~ parameter, scales = "free_y", ncol = 2) +
  labs(
    title = "Trace Plots for Selected Regression Coefficients",
    x = "MCMC Iteration",
    y = "Posterior Draw"
  ) +
  theme_minimal(base_size = 13)
```

## 7.3 Trace plot for residual variance
```{r }
sigma_df <- tibble(
  iteration = 1:length(sigma2_post),
  sigma2 = sigma2_post
)

sigma_df %>%
  filter(iteration %% 5 == 0) %>%
  ggplot(aes(x = iteration, y = sigma2)) +
  geom_line(linewidth = 0.3, alpha = 0.7) +
  labs(
    title = "Trace Plot for Residual Variance",
    x = "MCMC Iteration",
    y = expression(sigma^2)
  ) +
  theme_minimal(base_size = 13)
```

## 7.4 Posterior density plots

Density plots summarize uncertainty in selected coefficients. 
```{r }
beta_long %>%
  filter(parameter %in% selected_params) %>%
  ggplot(aes(x = value)) +
  geom_density(alpha = 0.4, fill = "steelblue") +
  facet_wrap(~ parameter, scales = "free", ncol = 2) +
  labs(
    title = "Posterior Density Plots for Selected Coefficients",
    x = "Coefficient Value",
    y = "Density"
  ) +
  theme_minimal(base_size = 13)
```


# 8. Posterior Summaries

We summarize posterior means, standard deviations, and 95% credible intervals.
```{r }
beta_summary <- apply(beta_post, 2, function(x) {
  c(
    mean = mean(x),
    sd = sd(x),
    q025 = unname(quantile(x, 0.025)),
    q975 = unname(quantile(x, 0.975))
  )
}) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("parameter") %>%
  mutate(
    parameter = recode(
      parameter,
      "(Intercept)" = "Intercept",
      "milage_std" = "Mileage (standardized)",
      "model_year_std" = "Model Year (standardized)",
      "accident_bin" = "Accident History",
      "trans_auto" = "Automatic Transmission",
      "fuel_typeDiesel" = "Fuel: Diesel",
      "fuel_typeElectric" = "Fuel: Electric",
      "fuel_typeGasoline" = "Fuel: Gasoline",
      "fuel_typeHybrid" = "Fuel: Hybrid",
      "fuel_typePlug-In Hybrid" = "Fuel: Plug-In Hybrid"
    )
  )

beta_summary
```

## 8.1 Coefficient plot with 95% credible intervals

This plot shows posterior means and corresponding credible intervals.
```{r }
beta_summary %>%
  mutate(parameter = reorder(parameter, mean)) %>%
  ggplot(aes(x = mean, y = parameter)) +
  geom_point() +
  geom_errorbarh(aes(xmin = q025, xmax = q975), height = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    title = "Posterior Means and 95% Credible Intervals",
    x = "Posterior Estimate",
    y = "Parameter"
  ) +
  theme_minimal(base_size = 13)
```


# 9. Posterior Predictive Analysis

We generate posterior predictive draws for the observed vehicles. 
```{r }
y_rep <- matrix(NA, nrow = nrow(beta_post), ncol = n)

for (s in 1:nrow(beta_post)) {
  mu <- X %*% beta_post[s, ]
  y_rep[s, ] <- rnorm(n, mean = mu, sd = sqrt(sigma2_post[s]))
}

pred_log <- colMeans(y_rep)

pred_price <- exp(pred_log)
obs_price <- exp(y)
```

## 9.1 Observed vs predicted prices
```{r }
pred_df <- tibble(
  observed_price = obs_price,
  predicted_price = pred_price
)

ggplot(pred_df, aes(x = observed_price, y = predicted_price)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(
    title = "Observed vs Predicted Prices",
    x = "Observed Price",
    y = "Predicted Price"
  ) +
  theme_minimal(base_size = 13)
```

## 9.2 Observed and predicted log-price distributions
```{r }
dist_df <- tibble(
  observed_log_price = y,
  predicted_log_price = pred_log
) %>%
  pivot_longer(
    cols = everything(),
    names_to = "type",
    values_to = "log_price"
  ) %>%
  mutate(
    type = recode(
      type,
      "observed_log_price" = "Observed",
      "predicted_log_price" = "Predicted"
    )
  )

ggplot(dist_df, aes(x = log_price, fill = type)) +
  geom_density(alpha = 0.4) +
  labs(
    title = "Observed and Predicted Log-Price Distributions",
    x = "Log Price",
    y = "Density"
  ) +
  theme_minimal(base_size = 13)
```



```{r }

```





