---
title: "Bayesian Model Analysis"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

# 1. Overview

This section implements the Bayesian regression model for used-car prices. 
Exploratory analysis and initial data cleaning were performed separately.

Goal: 

-   estimate the effect of vehicle characteristics on used-car price

-   quantify uncertainty in these effects using posterior distributions

-   generate predictive distributions for car prices

We model **log(price)** because raw prices are right-skewed.


# 2. Load packages 

```{r }
library(tidyverse)
library(janitor)
library(forcats)
library(MASS)
library(coda)
```


# 3. Load and prepare data

```{r }
cars <- read_csv("../data/train.csv") %>%
  clean_names()
```

We start from the cleaned dataset used in the exploratory analysis.

## 3.1 Restrict to Major Brands
```{r }
top_brands <- c(
  "BMW","Ford","Mercedes-Benz","Chevrolet","Audi","Porsche",
  "Toyota","Lexus","Jeep","Land","Cadillac","Nissan"
)

cars_top <- cars %>%
  filter(brand %in% top_brands)
```

## 3.2 Create Modeling Variables

We construct simplified variables for regression.
```{r }
cars_model <- cars_top %>%
  mutate(
    log_price = log(price),

    accident_bin = if_else(grepl("None", accident),0,1),

    trans_auto = if_else(
      grepl("Automatic|A/T|CVT|Dual", transmission, ignore.case = TRUE),
      1,0
    ),

    milage_std = as.numeric(scale(milage)),
    model_year_std = as.numeric(scale(model_year))
  )
```

Continuous predictors are standardized to improve numerical stability.

## 3.3 Remove Missing Values

For simplicity, we use complete-case analysis. 
```{r }
cars_model <- cars_model %>%
  drop_na(
    log_price,
    milage_std,
    model_year_std,
    accident_bin,
    trans_auto,
    fuel_type,
    brand
  )
```


# 4. Construct the Design Matrix

We build the regression design matrix using selected predictors. 
```{r }
X <- model.matrix(
  ~ milage_std + model_year_std + accident_bin + trans_auto + fuel_type + brand,
  data = cars_model
)

y <- cars_model$log_price

n <- nrow(X)
p <- ncol(X)
```


# 5. Bayesian Model Setup 

Model: 
$$
y \sim N(X\beta,\sigma^2)
$$
Priors:

-   $\beta \sim N(0,\tau^2 I)$

-   $\sigma^2 \sim IG(a,b)$ 

```{r }
tau2 <- 100^2
a0 <- 0.01
b0 <- 0.01
```

These are weakly informative priors.


# 6. Gibbs Sampling

We implement Gibbs sampling using the conjugate full conditional distributions.
```{r }
set.seed(724)

n_iter <- 12000
burn <- 2000

beta_draws <- matrix(NA,n_iter,p)
sigma2_draws <- numeric(n_iter)

beta <- rep(0,p)
sigma2 <- var(y)

XtX <- t(X)%*%X
Xty <- t(X)%*%y
Ip <- diag(p)

for(i in 1:n_iter){

  Vb <- solve(XtX/sigma2 + Ip/tau2)
  mb <- Vb %*% (Xty/sigma2)

  beta <- as.numeric(MASS::mvrnorm(1,mb,Vb))

  resid <- y - X%*%beta

  a_n <- a0 + n/2
  b_n <- b0 + sum(resid^2)/2

  sigma2 <- 1/rgamma(1,a_n,b_n)

  beta_draws[i,] <- beta
  sigma2_draws[i] <- sigma2
}

beta_post <- beta_draws[(burn+1):n_iter,]
sigma2_post <- sigma2_draws[(burn+1):n_iter]

colnames(beta_post) <- colnames(X)
```


# 7. MCMC Diagnostics



```{r }

```



```{r }

```



```{r }

```



```{r }

```



```{r }

```



```{r }

```



```{r }

```



```{r }

```



```{r }

```



```{r }

```



```{r }

```



```{r }

```



```{r }

```



```{r }

```



```{r }

```



```{r }

```



```{r }

```



```{r }

```



```{r }

```



```{r }

```



```{r }

```



```{r }

```



```{r }

```



```{r }

```
